---
title: "Homework 04"
author: "Spencer Pease"
date: "10/28/2019"
output: pdf_document
---

```{r setup, include=FALSE}

# Set output options
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')


# Include libraries
library(magrittr)
library(ggplot2)

# Source helper functions
source("functions/summary_table.R")
source("functions/population_estimates.R")


# Load data
mri <-
  read.table("data/mri.txt", header = TRUE) %>%
  dplyr::as_tibble() %>%
  dplyr::select(
    .data$packyrs, .data$crt, .data$male, .data$obstime, .data$death
  )

mri_surv <- mri %>%
  dplyr::mutate(
    `Survival Status` = dplyr::if_else(
      .data$obstime >= 5 * 365,
      "5+ years", "<5 years"
    )
  ) %>%
  dplyr::group_by(.data$`Survival Status`)


```


# Overview of the Data

The data in this report comes from a study of the risk factors associated with
cardiovascular and cerebrovascular disease among generally healthier older
adults. Specifically, this study enrolled a cohort of older (65+) and generally
healthy adults by randomly sampling individuals enrolled in the United States
Medicare system. Overall **`r nrow(mri)`** adults agreed to participate in the
study.

This report focuses on five variables from the study data. They are briefly
described below (see the study documentation for more details):

```{r variable-description, fig.cap="Descriptions of relevant variables",}

var_desc <- tibble::tribble(
  ~Variable, ~Units, ~Description,
  "packyrs", "pack years (1 pack of cigarettes per day for 1 year)", "Smoking history",
  "crt", "mg/dl", "Measure of creatinine in the participant's blood",
  "male", "Binary (0 = female, 1 = male)", "Binary Indicator of patient sex",
  "obstime", "days", "Total time the participant was observed on the study",
  "death", "binary", "Indicator that the participant died during the study"
)

knitr::kable(var_desc, caption = "Variable Descriptions")

```
